Detection and replication of epistasis influencing transcription in humans

### 2013-06-11 docs/manuscript/nature_letter.tex
changed title
awaiting final graphs from kostya

\title{Epistasis is widespread in the genetic control of transcription in humans}
\date{}
\author{Gibran Hemani, Konstantin Shakhbazov, Joseph E Powell}

@@

\begin{abstract}
A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Carlborg2004, Hill2008a, Crow2010}. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, evidence for multiple instances of epistasis is found. In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 238 genes. Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing. Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$). We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p < 1.8 \times 10^{-10}$). Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs. For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first strong evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans, and demonstrates its significant contribution to the variance of gene expression.
\end{abstract}

@@

However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome level \cite{Visscher2008} but this is intractable for non-additive variance components. Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes. By contrast, the epistatic signals detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance. This is broadly in line with theory \cite{Hill2008a}, but there are two caveats to this comparison. Firstly the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. Secondly the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs. Therefore this is likely to be a lower bound on the estimate of the relative contribution of epistasis to the genetic variance of gene expression.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{graph_of_interactions.pdf}
	\caption{\textbf{Replication of epistatic networks in two cohorts} Panel A represents all 434 genetic interactions (edges) discovered in the discovery set. Panel B shows those interactions that were present at the 5\% FDR level in at least one replication dataset, and panel C shows which interactions were significant in both replication datasets. Red nodes represent probes and black nodes represent SNPs. The congruence between datasets is indicitive of the consistent epistatic instances in independent populations.}
	\label{fig:fireworks}
\end{figure}
\clearpage

>>>>

\title{Detection and replication of epistasis influencing transcription in humans}
\date{}
\author{Gibran Hemani, Konstantin Shakhbazov, Joseph E Powell}

@@

\begin{abstract}
A long standing question in evolution and human genetics is the extent to which epistasis, the phenomenon whereby one polymorphism's effect on a trait depends on other polymorphisms present in the genome, contributes to complex traits \cite{Carlborg2004, Hill2008a, Crow2010}. Though epistasis has been demonstrated in artificial gene manipulation studies in model organisms \cite{Costanzo2010, Bloom2013}, and some examples have been shown in other species \cite{Carlborg2006}, few convincing examples exist for epistasis amongst natural polymorphisms in human traits \cite{Strange2010, Evans2011}. Its absence from empirical findings may simply be due to its unimportance in the genetic control of complex traits \cite{Hill2008a, Crow2010}, but we sought to test the hypothesis that it has previously been too technically difficult to detect due to statistical power and computational issues \cite{Cordell2009}. Here we show that, using advanced computation techniques \cite{Hemani2011} and a gene expression study design, evidence for multiple instances of epistasis is found. In a cohort of 846 individuals with data on 7339 gene expression levels in whole blood, we found that after stringent correction for multiple testing there were 501 significant pairwise epistatic interactions acting on the expression levels of 238 genes. Thirty interactions replicated in two independent datasets \cite{Metspalu2004, Fehrmann2011} following Bonferroni correction for multiple testing. Of the SNPs that did not pass this replication threshold, there was significant enrichment for interaction effects (316 of 434 above the 2.5\% confidence interval for uniformly distributed $p$-values, $p < 1.0 \times 10^{-16}$). We provide evidence of functional enrichment for the interacting SNPs, for instance 44 of the genetic interactions are located within 5Mb of regions of known intra-cellular chromosome interactions \cite{Lieberman-Aiden2009} ($p = 1.8 \times 10^{-10}$). Multi-locus epistatic interactions control 129 genes, whereby one \emph{cis}-acting single nucleotide polymorphism (SNP) is modulated by several trans-acting SNPs. For example MBNL1 is controlled by an additive effect at rs13069559 which itself is controlled by \emph{trans}-SNPs on 14 different chromosomes, with nearly identical genotype-phenotype (GP) maps for each \emph{cis}-\emph{trans} interaction. This study presents the first strong evidence for multiple instances of epistatic genetic effects emerging from natural genetic variation in humans, and demonstrates its significant contribution to the variance of gene expression.
\end{abstract}

@@

However in terms of their contribution to complex traits a more important metric might be the proportion of the variance that the epistatic loci explain \cite{Hill2008a}. Ideally one would approach this question from a whole genome level \cite{Visscher2008} but this is intractable for non-additive variance components. Yet some inference can be made from the ascertained effects in these analyses and it is evident that additive variance is overall a larger component than epistatic variance, as has been argued previously \cite{Hill2008a, Crow2010}. Taking the additive effects detected in Powell \emph{et al} (2012) at the $p < 2.91 \times 10^{-16}$ threshold, we calculate that in total they explain 1.73\% of the total phenotypic variance of all 7339 probes. By contrast, the epistatic variance from the interacting SNPs detected in this study explain 0.25\% of all phenotypic variance, approximately seven times lower than the additive variance. This is broadly in line with theory \cite{Hill2008a}, but there are three caveats to this comparison. Firstly, the ratio of additive to epistatic variance may differ at different effect sizes. Secondly, the power of a 1 \emph{d.f.} test exceeds that of an 8 \emph{d.f.} test. And thirdly, the non-additive variance at causal variants is expected to be underestimated by observed SNPs in comparison to estimates for additive variance, due to differences in the rate of decay of the estimate of the genetic variance of the causal SNPs as LD decreases with the observed SNPs.

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=5in]{Grey_graph_of_interactions_2_lists.pdf}
	\caption{\textbf{Replication of epistatic networks in two independent cohorts} All 434 putative genetic interactions (edges) with data common to discovery and replication sets is shown. 316 Interactions that had $p$-values exceeding the 2.5\% confidence interval following meta analysis of the replication data are shown with black nodes for SNPs and red nodes for gene expression traits. The interactions that fell within the confidence interval and showed no evidence of replication are greyed out. There is strong congruence of interactions between independent populations.}
	\label{fig:fireworks}
\end{figure}
\clearpage
